---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_knit$set(root.dir = tempdir())
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

```

# RFI: R to Modern Fortran Interface

<!-- badges: start -->
<!-- badges: end -->


This R package provides `.ModernFortran()`, an interface similar to `.Fortran()`
but for Fortran 2018.

The 2018 Fortran language standard expanded support for interfacing C and Fortran. 
One of the additions is the introduction of *C descriptors*, a data structure 
for passing arrays between C and Fortran. This package takes advantage of this. 

In contrast with `.Fortran`, R arrays are not passed as naked pointers, but as
*C descriptors* that contain information about rank, shape, element size, type,
and memory stride of the array that the Fortran routine can access directly.
This means that arguments for passing the size or rank of arrays is no longer
needed, which should lead to cleaner, simpler Fortran code. Additionally,
logical types are now supported directly.

Currently supported type conversions are:

|R type    |Fortran type                |
|:---------|:---------------------------|
|`logical` |`logical(c_bool)`           |
|`integer` |`integer(c_int)`            |
|`double`  |`real(c_double)`            |
|`complex` |`complex(c_double_complex)` |


```{r, echo=FALSE, eval = FALSE, results='asis'}
knitr::kable(matrix(ncol = 2, byrow = TRUE,
  dimnames = list(NULL, c("R type", "Fortran type")),
  sprintf("`%s`",
    c(
      "logical", "logical(c_bool)",
      "integer", "integer(c_int)",
      "double", "real(c_double)",
      "complex", "complex(c_double_complex)"
    )
  )))
```

## Installation

You can install RFI from github:
```r
if(!requireNamespace("remotes")) install.packages("remote")
remotes::install_github("t-kalinowski/RFI")
```

## Example

Say you have a Fortran 2018 module with a subroutine like so:

```{r include=FALSE}
fmodule <- 
"module mod_cshift

   use iso_c_binding, only: c_int, c_double, c_double_complex, c_bool
   implicit none

contains

   subroutine my_subroutine(array, shift) bind(c)
      integer(c_int), intent(in out) :: array(:)
      integer(c_int), intent(in) :: shift

      array = cshift(array, shift)
   end subroutine my_subroutine

end module mod_cshift
"
```

```{r, results='asis', echo = FALSE}
writeLines(paste("```f90\n", fmodule, "```"))
```
(`cshift` by the way is Fortran intrinsic for **circular shift**)


```{r include=FALSE}
f_module_file <- "mod_my_fortran_module.f90"
tmpdir <- writeLines(fmodule, f_module_file)
```


From R you can compile it like so:
```{r}
f_module_file <- "mod_my_fortran_module.f90"
so_file <- sub("f90$", "so", f_module_file)

compile_cmd <- sprintf(
  "gfortran -std=f2018 -shared -lgfortran -lm -lquadmath %s -o %s",
  f_module_file, so_file)

system(compile_cmd)
```

(you could also use the official R pathway to creating shared objects), with something like: 

```{r, eval = FALSE}
compile_cmd <- sprintf(
  "PKG_FFLAGS=-std=f2018 R CMD SHLIB %s -o %s",
  f_module_file, so_file)
system(compile_cmd)
```

Once the shared object is made, use it from R like so:
```{r}
# load the fortran module
dll <- dyn.load(so_file)

# get C pointer to the Fortran subroutine C entry point
func_ptr <- getNativeSymbolInfo('my_subroutine', dll)$address

# call the C function with R arrays
RFI::.ModernFortran(func_ptr, array=1:5, shift=2L)
```

Just like the other interfaces to compiled code (`.Fortran`,`.Call`, etc),
you'll probably want to wrap this in an R function for convince and input type
coercion.
```{r}

cshift <- function(array, shift) {
  RFI::.ModernFortran(func_ptr, 
                      array = as.integer(array), 
                      shift = as.integer(shift))$array
}

cshift(1:10, -3)
```

Tested with gfortran 9.2 and 9.3, R 3.6.3 and R 4.0.

